### 2016-09-21 doc/publications/plos16/plos16.tex
added Piotr's table

%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
% >>CONDENSE
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\subsection{Overview of Random Forests}

@@

% <<CONDENSE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\begin{figure}[tbhp]
%\begin{tabular}{ll}
%a)& b)\\
%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
% \end{tabular}
% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
%   variable with added Gaussian noise (in red).}
%  \label{figure:synth}
%\end{figure}
% <<CONDENSE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
\hline
%\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
&&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
&&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
&&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
 \hline
phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
&&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
&&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
&&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
&&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
\hline
phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
&&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
&&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
&&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
  was done on dpas-03} \\ 
\hline
phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
&&& RF - Spark ML & - & -  & -\\
&&& LR - Spark ML & & 14hr 48min & 8GB \\ 
&&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
\hline
phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
&&& RF - Spark ML & - & - & - \\
&&& LR - Spark ML & - & - & - \\ 
%&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

>>>>

%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.  
%DENIS Rob do you want to flesh this out more
\marginpar{Rob do you want to flesh this out more}
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.

However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features. 
Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
Using VariantSpark, we successfully built a k-means model on the full
$2500 \times 80$ million matrix to cluster individuals by
their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

Here we extend this toolkit to encompass supervised learning using random forest. 
Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
Although Spark ML can build a RF model on
a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
Spark ML uses, which we elaborate on in the discussion.

We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
% Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
% excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
% the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do
  we need to spell this out better, it should be a result section.}  
Harnessing the virtually unlimited capability to parallelise
tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc
  r2VIM}~\cite{Szymczak2016} on smaller datasets.

=======
%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
% 
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\subsection{Overview of Random Forests}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\begin{figure}[tbhp]
%\begin{tabular}{ll}
%a)& b)\\
%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
% \end{tabular}
% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
%   variable with added Gaussian noise (in red).}
%  \label{figure:synth}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}                               \\ \hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest  & 0.077 (XX) &                  &                                                                   \\
                                                & RF - Spark ML &            &                  &                                                                   \\
                                                & RF - R        & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger -R     & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger c++    & 0.94       & 7 mins           & --nthreads 20                                                     \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                                     \\
                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)}                \\
                                                    & ranger  -R    & 0.94       & 4 1/2 mins       & 21GB (one processor)                                                                \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & CursedForest  & 0.063 (XX) & 1hr 58min        & 4GB                                                                                 \\
                                                    & RF - Spark ML & 0.94       & 8hr 29min        & 8GB                                                                                 \\
                                                    & LR - Spark ML &            & 2hr 6min         & 8GB                                                                                 \\ 
                                                    & ranger  R     & 0.95       & 28 mins          & num.threads=10\footnote{don't
                                                                                                      know how to get the memory
                                                                                                      used,   This was done on dpas-03}  \\ 
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & CursedForest  & 0.046 (XX) & 4hr 0min                                             & 4GB         \\
                                                    & RF - Spark ML &            &                                                      &             \\
                                                    & LR - Spark ML &            & 14hr 48min                                           & 8GB         \\ 
                                                    & ranger  R     &  0.97      & 58 mins\footnote{ but took 6 hours to load the
                                                                                   data}
                                                                                                                                         &
                                                                                                                                           909GB\footnote{I
                                                                                                                                           think
                                                                                                                                           this
                                                                                                                                           is
                                                                                                                                           total
                                                                                                                                           memory. The
                                                                                                                                           job
                                                                                                                                           was
                                                                                                                                           run
                                                                                                                                           with
                                                                                                                                                   ntasks-per-node=160
                                                                                                                                           cores-per-socket=10
                                                                                                                                           on
                                                                                                                                           ruby
                                                                                                                                           }  \\ 
& ranger   -R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
& ranger (C++)       &        NA     &        NA     &             80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA      &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\

% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
% \hline
% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
%  \hline
% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
% \hline
% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
%   was done on dpas-03} \\ 
% \hline
% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
% &&& RF - Spark ML & - & -  & -\\
% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
% \hline
% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
% &&& RF - Spark ML & - & - & - \\
% &&& LR - Spark ML & - & - & - \\ 
% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
% \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

\begin{table}[!ht]
\centering
\caption{
{\bf Time in seconds to build random forest with 100
  trees with specified mtry fraction using given number of CPU cores.}}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
                                                            &  & 16   & 32      & 64      & 128  \\
\multirow{3}{*}{mtry (fraction)}                     & 0.10 & 708.27  & 340.00  & 186.01 & 134.47 \\	
                                                     & 0.25 & 1134.89 & 535.01  & 270.78 & 179.59 \\	
                                                     & 0.80 & 2313.47 & 1125.70 & 569.10 & 346.58 \\	
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{table10}
\end{table}
